---
title: "Sepsis Risk Prediction - Thesis Analysis Notebook"
subtitle: "ML and Classical Statistical Models vs. SIRS-Based Benchmark"
author: "Thesis Project"
date: "`r Sys.Date()`"

output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: show

# output:
#   github_document:
#     toc: true
#     toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = TRUE,
  warning = FALSE,
  cache   = TRUE
)
```

---

> **Research question:** How well does a machine-learning and classical statistical model
> trained on sepsis-related data predict the risk of sepsis compared to a SIRS-based
> benchmark, when evaluated in terms of AUC, Brier score, calibration, and decision-curve
> analysis using the PhysioNet Sepsis Challenge dataset?

---

# 1. Libraries

> - `tidyverse` / `tidymodels` — data wrangling and modelling framework
> - `slider` — rolling window calculations for feature engineering
> - `ggcorrplot` — correlation heatmap
> - `riskRegression` — model comparison with confidence intervals and contrasts
> - `probably` — calibration plots
> - `dcurves` — decision curve analysis
> - `shapviz` — SHAP feature importance for XGBoost

```{r libraries, cache=FALSE}
library(tidyverse)
library(tidymodels)
library(slider)
library(ggcorrplot)
library(riskRegression)
library(probably)
library(dcurves)   
library(shapviz)
library(ggplot2)

tidymodels_prefer()
```

---

# 2. Load Data

> - `sepsis_data` — full dataset, used only for EDA in Section 3
> - `train` / `test` — Sets used for all modelling, after split

```{r load-data, cache=FALSE}
sepsis_data <- read_csv("/home/william/Documents/R-markdown files/sepsis_data.csv")
train       <- read_csv("/home/william/Documents/R-markdown files/train.csv")
test        <- read_csv("/home/william/Documents/R-markdown files/test.csv")
```

---

# 3. Exploratory: Correlation & Missingness

> - Explores the full dataset before any modelling
> - Informs variable exclusion and preprocessing decisions

## 3.1 Missing Data Summary
> - Percentage of missing values per variable, sorted descending

```{r missing-summary, cache=TRUE}
missing_summary <- sepsis_data %>%
  summarize(across(everything(), ~ mean(is.na(.x)) * 100)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing") %>%
  arrange(desc(pct_missing))

missing_summary
```

## 3.2 Pairwise Correlation (training numeric predictors)

> - Spearman correlation matrix on all numeric predictors
> - Variables with |r| > 0.8 are removed downstream by `step_corr`

```{r correlation-matrix, cache=TRUE}
cor_matrix <- sepsis_data %>%
  select(where(is.numeric), -SepsisLabel) %>%
  cor(use = "pairwise.complete.obs", method = "spearman")
```

```{r correlation-heatmap, cache=TRUE, fig.width=11, fig.height=9}
ggcorrplot(cor_matrix,
           hc.order = TRUE, type = "full",
           outline.color = "grey",
           ggtheme = ggplot2::theme_gray,
           colors = c("#6D9EC1", "white", "#E46726")
)
```


## 3.3 % of patients with at least one measurement
```{r patient coverage, cache=TRUE}
#made with ChatGPT

patient_coverage <- function(df, id_col = "patient_id") {
  stopifnot(id_col %in% names(df))
df %>%
  mutate(across(-all_of(id_col), as.character)) %>% 
  select(all_of(id_col), everything()) %>%
  pivot_longer(
    cols = -all_of(id_col),
    names_to = "variable",
    values_to = "value"
  ) %>%
  group_by(variable) %>%
  summarise(
    patients_with_any_value = n_distinct(.data[[id_col]][!is.na(value)]),
    total_patients = n_distinct(.data[[id_col]]),
    pct_patients_with_any_value = (patients_with_any_value / total_patients)*100,
    .groups = "drop"
  ) %>%
  arrange(desc(pct_patients_with_any_value), desc(patients_with_any_value))
}

patient_coverage(sepsis_data, id_col = "patient_id")

```

---

# 4. Shared Preprocessing Functions (LogReg & XGBoost)

> - Excluded variables — sparse or clinically redundant features removed
> - Missingness indicators — binary flags capturing which values were missing before imputation
> - Rolling observation counts — measurement availability over 6h and 12h windows
> - Forward-fill — carries last known value forward within each patient's time series
> - Rolling means and SDs — captures short-term physiological trends over 6h and 12h

```{r shared-functions, cache=FALSE}
# Variables excluded from the ML models
excluded_variables <- c(
  "DBP", "TroponinI", "EtCO2", "PaCO2", "SaO2", "BaseExcess",
  "HCO3", "Hct", "AST", "Alkalinephos", "Bilirubin_direct", "Bilirubin_total",
  "PTT", "Fibrinogen", "Unit1", "Unit2"
)

# Forward-fill within each patient's time series
ffill_dataset <- function(df) {
  df %>%
    arrange(patient_id, ICULOS) %>%
    group_by(patient_id) %>%
    fill(-patient_id, -ICULOS, .direction = "down") %>%
    ungroup()
}

# Binary missingness indicators (1 = was NA before forward-fill)
add_indicator_columns <- function(df, outcome = "SepsisLabel") {
  No_indicator_variables <- c("patient_id", "ICULOS", outcome)
  df %>%
    mutate(across(
      -any_of(No_indicator_variables),
      ~ as.integer(is.na(.x)),
      .names = "was_na_{.col}"
    ))
}

# Rolling observation counts (6h / 12h windows)
add_obs_count_features <- function(df) {
  df %>%
    arrange(patient_id, ICULOS) %>%
    group_by(patient_id) %>%
    mutate(
      HR_obs_6h          = slide_int(HR,         ~sum(!is.na(.x)), .before = 5,  .complete = FALSE),
      Temp_obs_6h        = slide_int(Temp,       ~sum(!is.na(.x)), .before = 5,  .complete = FALSE),
      Resp_obs_6h        = slide_int(Resp,       ~sum(!is.na(.x)), .before = 5,  .complete = FALSE),
      MAP_obs_6h         = slide_int(MAP,        ~sum(!is.na(.x)), .before = 5,  .complete = FALSE),
      SBP_obs_6h         = slide_int(SBP,        ~sum(!is.na(.x)), .before = 5,  .complete = FALSE),
      WBC_obs_6h         = slide_int(WBC,        ~sum(!is.na(.x)), .before = 5,  .complete = FALSE),
      FiO2_obs_6h        = slide_int(FiO2,       ~sum(!is.na(.x)), .before = 5,  .complete = FALSE),
      Lactate_obs_6h     = slide_int(Lactate,    ~sum(!is.na(.x)), .before = 5,  .complete = FALSE),
      Platelets_obs_6h   = slide_int(Platelets,  ~sum(!is.na(.x)), .before = 5,  .complete = FALSE),
      Creatinine_obs_6h  = slide_int(Creatinine, ~sum(!is.na(.x)), .before = 5,  .complete = FALSE),
      HR_obs_12h         = slide_int(HR,         ~sum(!is.na(.x)), .before = 11, .complete = FALSE),
      Temp_obs_12h       = slide_int(Temp,       ~sum(!is.na(.x)), .before = 11, .complete = FALSE),
      Resp_obs_12h       = slide_int(Resp,       ~sum(!is.na(.x)), .before = 11, .complete = FALSE),
      MAP_obs_12h        = slide_int(MAP,        ~sum(!is.na(.x)), .before = 11, .complete = FALSE),
      SBP_obs_12h        = slide_int(SBP,        ~sum(!is.na(.x)), .before = 11, .complete = FALSE),
      WBC_obs_12h        = slide_int(WBC,        ~sum(!is.na(.x)), .before = 11, .complete = FALSE),
      FiO2_obs_12h       = slide_int(FiO2,       ~sum(!is.na(.x)), .before = 11, .complete = FALSE),
      Lactate_obs_12h    = slide_int(Lactate,    ~sum(!is.na(.x)), .before = 11, .complete = FALSE),
      total_obs_6h       = HR_obs_6h + Temp_obs_6h + Resp_obs_6h + MAP_obs_6h +
                           SBP_obs_6h + WBC_obs_6h + FiO2_obs_6h + Lactate_obs_6h +
                           Platelets_obs_6h + Creatinine_obs_6h
    ) %>%
    ungroup()
}

# Rolling means (6h + 12h) and rolling SDs (6h)
add_rolling_features <- function(df) {
  df %>%
    arrange(patient_id, ICULOS) %>%
    group_by(patient_id) %>%
    mutate(
      # 6h rolling means
      HR_roll_mean_6         = slide_dbl(HR,         ~mean(.x, na.rm = TRUE), .before = 5,  .complete = FALSE),
      Temp_roll_mean_6       = slide_dbl(Temp,       ~mean(.x, na.rm = TRUE), .before = 5,  .complete = FALSE),
      Resp_roll_mean_6       = slide_dbl(Resp,       ~mean(.x, na.rm = TRUE), .before = 5,  .complete = FALSE),
      MAP_roll_mean_6        = slide_dbl(MAP,        ~mean(.x, na.rm = TRUE), .before = 5,  .complete = FALSE),
      SBP_roll_mean_6        = slide_dbl(SBP,        ~mean(.x, na.rm = TRUE), .before = 5,  .complete = FALSE),
      WBC_roll_mean_6        = slide_dbl(WBC,        ~mean(.x, na.rm = TRUE), .before = 5,  .complete = FALSE),
      FiO2_roll_mean_6       = slide_dbl(FiO2,       ~mean(.x, na.rm = TRUE), .before = 5,  .complete = FALSE),
      Lactate_roll_mean_6    = slide_dbl(Lactate,    ~mean(.x, na.rm = TRUE), .before = 5,  .complete = FALSE),
      Platelets_roll_mean_6  = slide_dbl(Platelets,  ~mean(.x, na.rm = TRUE), .before = 5,  .complete = FALSE),
      Creatinine_roll_mean_6 = slide_dbl(Creatinine, ~mean(.x, na.rm = TRUE), .before = 5,  .complete = FALSE),
      # 12h rolling means
      HR_roll_mean_12        = slide_dbl(HR,         ~mean(.x, na.rm = TRUE), .before = 11, .complete = FALSE),
      Temp_roll_mean_12      = slide_dbl(Temp,       ~mean(.x, na.rm = TRUE), .before = 11, .complete = FALSE),
      Resp_roll_mean_12      = slide_dbl(Resp,       ~mean(.x, na.rm = TRUE), .before = 11, .complete = FALSE),
      MAP_roll_mean_12       = slide_dbl(MAP,        ~mean(.x, na.rm = TRUE), .before = 11, .complete = FALSE),
      SBP_roll_mean_12       = slide_dbl(SBP,        ~mean(.x, na.rm = TRUE), .before = 11, .complete = FALSE),
      WBC_roll_mean_12       = slide_dbl(WBC,        ~mean(.x, na.rm = TRUE), .before = 11, .complete = FALSE),
      FiO2_roll_mean_12      = slide_dbl(FiO2,       ~mean(.x, na.rm = TRUE), .before = 11, .complete = FALSE),
      Lactate_roll_mean_12   = slide_dbl(Lactate,    ~mean(.x, na.rm = TRUE), .before = 11, .complete = FALSE),
      # 6h rolling SDs
      HR_roll_sd_6           = slide_dbl(HR,         ~sd(.x, na.rm = TRUE), .before = 5, .complete = FALSE),
      Temp_roll_sd_6         = slide_dbl(Temp,       ~sd(.x, na.rm = TRUE), .before = 5, .complete = FALSE),
      Resp_roll_sd_6         = slide_dbl(Resp,       ~sd(.x, na.rm = TRUE), .before = 5, .complete = FALSE),
      MAP_roll_sd_6          = slide_dbl(MAP,        ~sd(.x, na.rm = TRUE), .before = 5, .complete = FALSE),
      SBP_roll_sd_6          = slide_dbl(SBP,        ~sd(.x, na.rm = TRUE), .before = 5, .complete = FALSE),
      WBC_roll_sd_6          = slide_dbl(WBC,        ~sd(.x, na.rm = TRUE), .before = 5, .complete = FALSE),
      FiO2_roll_sd_6         = slide_dbl(FiO2,       ~sd(.x, na.rm = TRUE), .before = 5, .complete = FALSE),
      Lactate_roll_sd_6      = slide_dbl(Lactate,    ~sd(.x, na.rm = TRUE), .before = 5, .complete = FALSE),
      Platelets_roll_sd_6    = slide_dbl(Platelets,  ~sd(.x, na.rm = TRUE), .before = 5, .complete = FALSE),
      Creatinine_roll_sd_6   = slide_dbl(Creatinine, ~sd(.x, na.rm = TRUE), .before = 5, .complete = FALSE)
    ) %>%
    ungroup()
}
```

## 4.1 Apply Preprocessing (Shared Train & Test)

```{r shared-preprocess, cache=TRUE}
train_preprocess_ffill <- train %>%
  select(-any_of(excluded_variables)) %>%
  add_indicator_columns(outcome = "SepsisLabel") %>%
  add_obs_count_features() %>%
  ffill_dataset() %>%
  add_rolling_features() %>%
  mutate(
    SepsisLabel = as.factor(SepsisLabel),
    Gender      = as.factor(Gender)
  )

test_preprocess_ffill <- test %>%
  select(-any_of(excluded_variables)) %>%
  add_indicator_columns(outcome = "SepsisLabel") %>%
  add_obs_count_features() %>%
  ffill_dataset() %>%
  add_rolling_features() %>%
  mutate(
    SepsisLabel = as.factor(SepsisLabel),
    Gender      = as.factor(Gender)
  )

cat("Preprocessed train rows:", nrow(train_preprocess_ffill), "\n")
cat("Preprocessed train columns:", ncol(train_preprocess_ffill), "\n")
cat("Preprocessed test rows:", nrow(test_preprocess_ffill), "\n")
```

---

# 5. Benchmark - SIRS Logistic Regression

> - SIRS criteria: HR > 90, Temp outside 36-38°C, RR > 20, WBC outside 4-12
> - Each met criterion adds 1, giving a score of 0-4
> - Score used as the sole predictor in a logistic regression
> - Represents traditional clinical sepsis screening practice

## 5.1 SIRS-Specific Preprocessing

```{r sirs-preprocess, cache=TRUE}
train_sirs <- train %>%
  select(-any_of(excluded_variables)) %>%
  ffill_dataset() %>%
  mutate(SepsisLabel = as.factor(SepsisLabel))

test_sirs <- test %>%
  select(-any_of(excluded_variables)) %>%
  ffill_dataset() %>%
  mutate(SepsisLabel = as.factor(SepsisLabel))
```

## 5.2 SIRS Recipe & Model

```{r sirs-recipe, cache=FALSE, dependson="sirs-preprocess"}
SIRS_recipe <- recipe(SepsisLabel ~ HR + Temp + Resp + WBC, data = train_sirs) %>%
  step_impute_mean(all_numeric_predictors()) %>%
  step_mutate(
    SIRS_score = (HR > 90) + (Temp > 38 | Temp < 36) + (Resp > 20) + (WBC > 12 | WBC < 4)
  ) %>%
  step_select(SepsisLabel, SIRS_score)

logistic_sirs_model <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")
```

## 5.3 Prep, Bake & Fit

```{r sirs-fit, cache=FALSE, dependson="sirs-recipe"}
Sirs_benchmark_prep          <- SIRS_recipe %>% prep(training = train_sirs)
Sirs_benchmark_training_bake <- Sirs_benchmark_prep %>% bake(new_data = NULL)
Sirs_benchmark_test_bake     <- Sirs_benchmark_prep %>% bake(new_data = test_sirs)

Logistic_sirs_fit <- logistic_sirs_model %>%
  fit(SepsisLabel ~ SIRS_score, data = Sirs_benchmark_training_bake)

Logistic_sirs_fit
```

## 5.4 Predictions & Evaluation

```{r sirs-eval, cache=FALSE, dependson="sirs-fit"}
prob_preds_sirs <- predict(Logistic_sirs_fit, new_data = Sirs_benchmark_test_bake, type = "prob")

sirs_results <- test_sirs %>%
  select(SepsisLabel) %>%
  bind_cols(prob_preds_sirs) %>%
  mutate(SepsisLabel = factor(SepsisLabel, levels = c("1", "0")))

test_sirs <- test_sirs %>%
  mutate(SIRS_Preds = prob_preds_sirs$.pred_1 * 100)

sirs_metrics <- metric_set(roc_auc, brier_class)(
  sirs_results, truth = SepsisLabel, .pred_1, event_level = "first"
)

sirs_metrics
```

---

# 6. Logistic Regression

> - Uses the full shared feature set including rolling features and missingness indicators
> - SIRS score added as an engineered predictor
> - Correlated features removed at 0.8 threshold
> - Interpretable ML baseline for comparison against XGBoost

## 6.1 Recipe & Model

```{r logreg-recipe, cache=FALSE, dependson="shared-preprocess"}
sepsis_recipe_log <- recipe(SepsisLabel ~ ., data = train_preprocess_ffill) %>%
  step_rm(patient_id) %>%
  step_dummy(Gender) %>%
  step_impute_mean(all_numeric_predictors()) %>%
  step_mutate(
    SIRS_score = (HR > 90) + (Temp > 38 | Temp < 36) + (Resp > 20) + (WBC > 12 | WBC < 4)
  ) %>%
  step_corr(threshold = 0.8)

logistic_sepsis_model <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")
```

## 6.2 Prep, Bake & Fit

```{r logreg-fit, cache=FALSE}
Sepsis_log_prep          <- sepsis_recipe_log %>% prep(training = train_preprocess_ffill)
Sepsis_log_training_bake <- Sepsis_log_prep %>% bake(new_data = NULL)
Sepsis_log_test_bake     <- Sepsis_log_prep %>% bake(new_data = test_preprocess_ffill)

Logistic_sepsis_fit <- logistic_sepsis_model %>%
  fit(SepsisLabel ~ ., data = Sepsis_log_training_bake)

Logistic_sepsis_fit
```

## 6.3 Predictions & Evaluation

```{r logreg-eval, cache=FALSE}
prob_preds_sepsis <- predict(Logistic_sepsis_fit, new_data = Sepsis_log_test_bake, type = "prob")

sepsis_log_results <- test_preprocess_ffill %>%
  select(SepsisLabel) %>%
  bind_cols(prob_preds_sepsis) %>%
  mutate(SepsisLabel = factor(SepsisLabel, levels = c("1", "0")))

log_reg_metrics <- metric_set(roc_auc, brier_class)(
  sepsis_log_results, truth = SepsisLabel, .pred_1, event_level = "first"
)

log_reg_metrics
```

---

# 7. XGBoost

> - Uses the same feature set as the logistic regression
> - Hyperparameters from prior tuning: 500 trees, depth 7, learning rate 0.012, early stopping at 20 rounds
> - 10% internal validation split used for early stopping
> - Primary ML model — captures non-linear interactions the logistic regression cannot

## 7.1 Recipe & Model

```{r xgb-recipe, cache=FALSE, dependson="shared-preprocess"}
sepsis_recipe_xgb <- recipe(SepsisLabel ~ ., data = train_preprocess_ffill) %>%
  step_rm(patient_id) %>%
  step_dummy(Gender) %>%
  step_impute_mean(all_numeric_predictors()) %>%
  step_mutate(
    SIRS_score = (HR > 90) + (Temp > 38 | Temp < 36) + (Resp > 20) + (WBC > 12 | WBC < 4)
  ) %>%
  step_corr(threshold = 0.8)

xgb_sepsis_model <- boost_tree(
  trees      = 500,
  tree_depth = 7,
  learn_rate = 0.01209315,
  stop_iter  = 20
) %>%
  set_engine("xgboost", counts = FALSE, eval_metric = "auc", validation = 0.1) %>%
  set_mode("classification")
```

## 7.2 Workflow & Fit

```{r xgb-fit, cache=FALSE}
xgb_wf <- workflow() %>%
  add_recipe(sepsis_recipe_xgb) %>%
  add_model(xgb_sepsis_model)

set.seed(123)

xgb_final_fit <- xgb_wf %>%
  fit(data = train_preprocess_ffill)

xgb_final_fit
```

## 7.3 Predictions & Evaluation

```{r xgb-eval, cache=FALSE}
prob_preds_xgb <- predict(xgb_final_fit, new_data = test_preprocess_ffill, type = "prob")

xgb_results <- test_preprocess_ffill %>%
  select(SepsisLabel) %>%
  bind_cols(prob_preds_xgb) %>%
  mutate(SepsisLabel = factor(SepsisLabel, levels = c("1", "0")))

xgb_metrics <- metric_set(roc_auc, brier_class)(
  xgb_results, truth = SepsisLabel, .pred_1, event_level = "first"
)

xgb_metrics
```

---

# 8. Model Comparison Summary

> - Formally compares all three models against the SIRS benchmark
> - AUC and Brier score reported with confidence intervals and pairwise contrast tests
> - IPA (Index of Prediction Accuracy) shows relative improvement over the null model

```{r comparison-table, cache=FALSE}
pred_list <- list(
  "#1 Benchmark (SIRS)"                = prob_preds_sirs$.pred_1,
  "#6 Logistic Regression"             = prob_preds_sepsis$.pred_1,
  "#10 XGBoost (tuned)"                = prob_preds_xgb$.pred_1
)


score_evaluation <- test %>%
  transmute(SepsisLabel = as.numeric(as.character(SepsisLabel)))

comparison <- Score(
  object    = pred_list,
  formula   = SepsisLabel ~ 1,
  data      = score_evaluation,
  metrics   = c("auc", "brier"),
  contrasts = TRUE,
  plots     = "calibration",
  summary   = "ipa"
)

summary(comparison)

cat("\n=== AUC contrasts (vs #1 Benchmark) ===\n");   print(comparison$AUC$contrasts)
cat("\n=== Brier contrasts (vs #1 Benchmark) ===\n"); print(comparison$Brier$contrasts)
cat("\n=== Brier + IPA ===\n"); print(comparison$Brier$score)
```

---

# 9. Calibration plots

> - Checks whether predicted probabilities match observed sepsis rates
> - A well-calibrated model follows the diagonal
> - Deviation indicates over- or under-estimation of risk

```{r calibration-plots, cache=FALSE}
cal_plot_windowed(sirs_results, truth = SepsisLabel, .pred_1,
                  event_level = "first", step_size = 0.025) + ggtitle("#1: SIRS Benchmark")

cal_plot_windowed(sepsis_log_results, truth = SepsisLabel, .pred_1,
                  event_level = "first", step_size = 0.025) + ggtitle("#6: Best Logistic Regression")

cal_plot_windowed(xgb_results, truth = SepsisLabel, .pred_1,
                  event_level = "first", step_size = 0.025) + ggtitle("#10 XGBoost")
```

---

# 10. Decision Curve Analysis (DCA)

> - Evaluates clinical utility across a range of decision thresholds (0-40%)
> - Net benefit weighs true positives against false positives at each threshold
> - A model is useful if its curve sits above both "treat all" and "treat none"

```{r Decision curves, cache=FALSE}
# DCA
dca_df <- tibble(
  SepsisLabel    = as.numeric(as.character(sirs_results$SepsisLabel)),
  SIRS_Benchmark = sirs_results$.pred_1,
  Best_LogReg    = sepsis_log_results$.pred_1,
  XGBoost        = xgb_results$.pred_1
)

dca(SepsisLabel ~ SIRS_Benchmark + Best_LogReg + XGBoost, data = dca_df, thresholds = seq(0, 0.40, by = 0.005)) %>%
  plot(smooth = TRUE) +
  labs(
    title = "Decision Curve Analysis – Three-Model Comparison",
    x     = "Threshold Probability",
    y     = "Net Benefit (per 100 patients)"
  ) +
  scale_x_continuous(breaks = seq(0, 0.40, by = 0.05), limits = c(0, 0.40), labels = scales::percent_format(accuracy = 1)) +
  scale_y_continuous(labels = function(x) x * 100) +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 13))
```

---

# 11. Feature Importance - SHAP values

> - SHAP values decompose each prediction into individual feature contributions
> - 16,000 test observations sampled for computational efficiency
> - Beeswarm plot shows both magnitude and direction of each feature's effect

```{r SHAP values, cache=FALSE}

set.seed(42)
xgb_test_baked <- extract_recipe(xgb_final_fit) %>%
  bake(new_data = test_preprocess_ffill %>% slice_sample(n = 16000)) %>%
  select(-SepsisLabel) %>%
  as.matrix()

shap_xgb <- shapviz(extract_fit_engine(xgb_final_fit), X_pred = xgb_test_baked)
shap_xgb$S <- -shap_xgb$S

feature_names <- c(
  ICULOS = "ICU Length of Stay", FiO2_obs_6h = "FiO2 Observations (6h)",
  HospAdmTime = "Hospital Admission Time", Temp = "Temperature",
  Creatinine = "Creatinine", SIRS_score = "SIRS Score",
  Temp_roll_mean_6 = "Temperature Rolling Mean (6h)", BUN = "Blood Urea Nitrogen",
  Resp_roll_mean_12 = "Respiratory Rate Rolling Mean (12h)", Lactate_obs_12h = "Lactate Observations (12h)",
  Lactate_obs_6h = "Lactate Observations (6h)", WBC = "White Blood Cell Count",
  Temp_obs_6h = "Temperature Observations (6h)", HR = "Heart Rate",
  HR_roll_mean_12 = "Heart Rate Rolling Mean (12h)"
)

safe_names <- feature_names[colnames(shap_xgb$S)]
safe_names[is.na(safe_names)] <- colnames(shap_xgb$S)[is.na(safe_names)]
colnames(shap_xgb$S) <- colnames(shap_xgb$X) <- safe_names

sv_importance(shap_xgb, kind = "beeswarm") + ggtitle("XGBoost – SHAP Feature Importance")
```


